### 2023-01-27 ms/ms.tex
revised ms

\title{\textbf{Asymmetries in Cricket \\ \vspace{5pt} \large{The Effects of Winning the Toss on Winning the Match}}\thanks{This paper incorporates some of the text and 
results from Sood \& Willis's (2019) working paper: ``Fairly Random: The Impact of Winning the Toss on the Probability of Winning.'' The authors thank Matt Blackwell, 
Andrew Gelman, Donald Green, Jens Hainmueller, and Daniel Stone for their comments and suggestions. We are grateful to Sajan Mehotra for outstanding research assistance. 
% The first three authors contributed roughly equally to the research design, data collection, and implementation of the whole project, and are listed alphabetically. Acharya
% contributed to the research design and interpretation of results.
}}

@@

%
\section{Introduction}
In a competitive sport, every little thing matters. Yet, many sports leave some large levers out of the reach of the teams and in the
hands of fate. In cricket, one such lever---the toss---has been the subject of much attention. In many matches, commentators and cricket enthusiasts claim that there is a definite advantage to either bowling or batting first.\footnote{See, for example, the \href{https://www.quora.com/How-much-does-a-toss-matter-to-a-game-of-cricket}{Quora page on ``How much does a toss matter to a game of cricket?''}} The advantage is also often pointed to by the captains in the pre-toss interview and by the captain of the losing team in the post-match interview. To wrest the advantage, the captain merely needs to pick the side of the coin that will be left facing the sky after the toss and choose whichever one---batting second or batting second---gives the greater advantage.

In this paper, we measure how well teams have been able to capitalize on any advantage conferred by winning the toss. Since the only consequence of the toss is letting the toss-winning side decide whether to bat or field first, we also attempt to measure the advantage of electing to bat first. To answer these questions, we collected data on over 35,000 cricket matches across four formats (First Class, ODIs, T20s, and tests) played over the course of more than one hundred fifty years. (To our knowledge, this is the most comprehensive professional cricket match dataset to date.)

Winning the toss increases the chances of winning the match by 1.8 percentage points: toss winners win 39.2\% of the time and lose 37.4\% of the time, with the remaining share being draws or ties. The effect of winning the toss is the largest in tests, where it is 4.9 percentage points and is smallest in ODIs and T20s, where it is essentially zero. That being said, in ODIs and T20s that are day/night matches, the effect of winning the toss is 3.3 percentage points, whereas, in the day-time matches, the effect is a precise zero.

The effect of toss in the ODIs has remained relatively stable since the 1980's when the one-day format first became popular. In tests, however, the importance of the toss has varied considerably. The effect of winning the toss in test cricket was a little over five percentage points before 1990. It then dropped nearly to 0 from 1990 to 2010. In the last decade, however, it has exploded, averaging 14.5 percentage points. Counter to conventional wisdom, the effect of winning the toss is relatively similar across continents, though the toss appears less important on European and African wickets than elsewhere. The effect on Aussie and Kiwi wickets (3 percentage points) is not substantially different from its effect on South Asian wickets (2.8 percentage points) but is higher than its effect on European and African wickets (1.1 and 0 percentage points respectively).

In sports like soccer, the conventional wisdom is that the toss is inconsequential \citep{csato2020comparison}. In cricket, it is believed that the toss has little significance in lopsided matches that have clear favorites and underdogs. For example, it is thought to be less relevant in an India-Ireland match than in an India-Australia match. For this reason, we look at the effect of the toss separately in subsamples where teams are similarly ranked according to ICC ELO ratings and where their ratings differ by more substantially. We find that the effect of the toss doesn't vary by how well matched the teams are. Since ELO ratings are available only for matches played in recent decades, these findings (along with our finding that influence of the toss has declined over time) suggest that the toss has little influence in modern cricket.

An important caveat about the null findings is in order, however. Our empirical approach cannot tell whether the effects are nil because the toss is unimportant or because the toss is important, but that (modern) teams have not been able to capitalize on the advantage. The toss may matter quite a bit in principle, but we would only be able to observe its effect if a team is able to use the information on the conditions of play and how it pairs strategically against the opposing team to be able to make the optimal decision of whether to bat or field first. If teams simply do not have this knowledge and are often choosing poorly, then we would not see any ``toss effects.'} So, when we refer to ``toss effects'' in this paper, what we are measuring is not whether there is an advantage to winning the toss in principle, but whether any such advantage is exploited in practice.

\footnote{The median rating is 17, so a cutoff of 15 roughly splits the sample of matches in which we have ratings for both teams in two halves.}

%Another way to benchmark the effect of the toss is to compute it in
%terms of ``run- or wicket-advantage.'' To do this, we consider ODI
%and T20 matches, and estimate [.Apoorva: describe the methodology
%that we discussed here]. We find that the effect of winning the toss
%and electing to bat first is an [X] run advantage. In other words,
%the effect of the toss in these instances is nullified only by
%allowing the chasing team to start their innings with [X] runs
%already on the board. Similarly, the effect of winning the toss and
%electing to field first is a [Y] run advantage, so that the effect is
%nullified only if we make the chasing team start with [-Y] runs on
%the board. \todo{We had an idea here that I vaguely recall of how to
%address the complication that if a team loses all ten wickets, then
%we don't observe its 50 over total.}

@@

However, we find that of the toss winners that elected to bat, only 38.1\% won the match compared to 37.9\% that lost--- a small difference of 0.2 percentage points. On the other hand, of the toss winners that elected to field, 40.8\% won the match compared to 36.9\% that lost--- a larger difference of 3.9 percentage points. This means that 94\% of the 1.8 percentage point effect of winning the toss can be attributed to instances in which the toss winning team elected to field, and only 6\% of the effect can be attributed to the toss winning team electing to bat. This suggests that teams are better able to capitalize on the toss by electing to field when it is advantageous to do so. 
%
%Combining this with the fact that we saw the reduced form effect of the toss on the outcome of the match to be 1.8 percentage points, we can uncover the average effect of batting first on the so-called ``complier'' teams: those that got to bat (because they won the toss) that would not have batted if batting first were randomly assigned \citep{angrist1996identification}. Since winning the toss results in a 0.9\% excess of match wins (from the benchmark case in which the toss has no effect), the effect of batting first in this complier population is 0.9/6.5 = 13.9 percentage points. This is the instrumented effect of winning the toss-- or, the rate at which the excess decisions to bat were converted into excess wins.\footnote{Since the choice to field first is the perfectly colinear inverse of the choice to bat first, flipping the sings of all of our estimates gives the instrumented effect of electing to field first. For example, the instrumented effect of electing to field first is -13.9 percentage points.} 
%
%This conversion rate is different in sign and considerably larger than the naive difference in win
%probabilities between teams that bat first and those that don't, which is $-1.6$ percentage points. Since the rules of cricket imply that the toss cannot affect the outcome of the match other than through the decision of the toss-winning team to bat or field, this discrepancy is likely due to the complier population being very different from the never-taker or always-taker populations that form the bulk of our team-match data. The IV estimates are thus
%telling us that some teams in some matches are able to take advantage of the decision to bat first when the batting choice really matters.

We also look at how the toss advantage breaks down between teams that elected to bat versus those that elected to field within the various subgroups mentioned above and others. Perhaps the most striking result from this analysis for many cricket followers is that there does not appear to be the case that toss winners that elect to bat first do better than those that elected to field first in rain delayed matches where the chasing total is adjusted according to the \citet{duckworth1998} formula.

Our results are pertinent to debates on how cricket should be governed. One argument is that since the toss is random, no team is advantaged by the format prior to the toss; and in fact the luck of the toss along with the strategic choice to bat first or second are integral parts of the game--- not very different from the luck of weather affecting the teams differently. 
%or the strategic choices that captains make as to how to set the field, or the order in which they deploy their batsmen.
An alternative argument, however, is that the toss is problematic because it advantages one team over the other even before the first ball is bowled; and care should be taken to mitigate this asymmetry--- for example, by making cricket pitches more durable so that they don't deteriorate so much so that batting on the final day of a test match is miserable. Otherwise, the toss-winning team could exploit this advantage by batting first, reducing the chance that it has to endure this misery. Moreover, the fact that no side is advantaged \emph{ex-ante} if the toss is random is not enough to justify not correcting any asymmetry created by the toss. In fact, the rules of cricket already try to correct the asymmetry between teams that bat first and those that bat second in one particular instance: In rain-delayed one-day matches, the Duckworth-Lewis formula is used to set a new target for the chasing team.\footnote{The argument that no side was advantaged \emph{ex-ante} could have been used to justify not doing any correction in such rain-delayed matches, and simply declaring that the team that batted first won the match because the chasing team was unable to reach its target. But the designers of one-day cricket were clearly unwilling to allow the bad luck of weather to have this much influence on the game. If addressing the misfortunes of weather are important, it may also be important to address the misfortune of losing the toss.}

@@

We do not address these optimal design questions in this paper, but our results are relevant to these issues depending on what the designers of cricket seek to optimize. %However, at the end of the paper, we propose (and discuss the possible consequences of) a subtle change to the rules governing the toss: Rather than tossing a coin to confer the choice to bat first or second upon the captain of the toss-winning team, umpires may consider simply tossing a coin to determine which team must bat first.

@@

We obtained data on cricket matches from the ESPN Cricinfo website, which provides coverage of international and many domestic matches,
and maintains statistics on a wide variety of matches across formats.\footnote{Information on all available matches was downloaded in June 2020 using a purpose-built parsing library available at \href{https://github.com/dwillis/python-espncricinfo}{https://github.com/dwillis/python-espncricinfo}. The data will be available as a separate repository available at \href{https://github.com/dwillis/cricket-stats}{https://github.com/dwillis/cricket-stats}.}

We focus on four types of formats: T20, ODI, Test, and First Class matches for both men's and women's teams.\footnote{We do not include the List A matches that are not ODIs or T20s, which are often considered a less professional category (with worse data recording) than the formats that we include.} The data are a near census of the relevant population, excluding only scheduled matches that were abandoned without the toss being conducted, and a small number of limited-overs matches in which the minimum number of overs that must be bowled to establish a result was not met.\footnote{In ODIs, for example, each side must bat at least 20 overs for a result to be declared. We have no reason to suspect that the small matches that were excluded because they did not meet this threshold could be biased enough to affect our conclusions. We drop 426 matches recorded as ``no result,'' 3 matches that are recorded as ``abandoned'' or ``canceled,'' and another three that don't record who bat first. The former constitutes the largest exclusion we make and appears to be comprised of matches canceled for idiosyncratic reasons, ranging from the \href{https://www.espncricinfo.com/series/shell-tri-series-1991-92-61203/australia-women-vs-england-women-final-66975/full-scorecard}{a team being unable to bat because of the weather} to cancellations because of \href{https://www.espncricinfo.com/series/australia-tour-of-pakistan-1982-83-61392/pakistan-vs-australia-3rd-odi-64196/full-scorecard}{unwanted audience participation}. This results in us dropping approximately 0.07\% of matches.}

Our final sample comprises of 35,357 cricket matches across four formats played over the last 150 years. We report the distribution of matches over time and by type in Figure \ref{fig:timeseries_decomp}. As the figure shows, the number of matches has exploded over the last two decades, largely due to the introduction of the enormously popular Twenty-20 (T20) format of the game both in both international and domestic tournaments. T20 matches constitute over a quarter of all matches played in the last decade.

@@

Since our analysis will be based on the assumption that the coin toss is random, we start by examining this assumption.\footnote{Prior work in the sports analytics of cricket by \citet{shafqat2015analysis} and \citet{dawson2009bat} conditions on post-treatment variables, whereas we use only the randomization of the toss to identify toss effects.} Is there any evidence that the coin toss is somehow rigged--- for example, with the home side enjoying the rub of the green more often? In order to study this, we compare the rates at which home teams, teams that won the toss in their last match, and underdogs (defined as the lower-ranked team whenever rankings are available for both teams) win the toss relative to their opponents.

To examine whether the toss is rigged, we compute the odds of seeing as many successes in the same number of draws of a sequence of independent Bernoulli random variables with parameter 0.5. For example, home teams have won 50.81\% of tosses in the 6377 matches for which we have data. The likelihood of tossing a fair coin 6377 times and getting at least this percentage of heads is 20.1\%. This chances appears low, but not eyebrow-raisingly so. We find that toss wins are even more balanced for teams that won and lost the last toss, and for teams that are underdogs and favorites. Table \ref{table: balance} reports these balance results.

That said, we do see one potential source of concern. We find that teams that share their nationality with at least two of the three umpires of the match win 51.3\% tosses in the 4044 international matches for which we have this data. The likelihood that tossing a fair coin 4044 times and and getting at least this share of heads is only 10\%. This number is certainly low enough to raise some eyebrows. That said, the practice of allowing home-team umpires in international matches ended in 1992, and so these issues are less relevant in recent decades which comprise the bulk of our sample.

In addition, a rigged toss in favor of the home team would bias estimates of the advantage of winning the coin toss to the extent that it is
correlated with ability. If stronger teams win more tosses, estimates of the advantage of the coin toss would be inflated upwards. And vice versa, if otherwise. We do not have good reasons to think that there is a correlation. So for now, we proceed as if the toss is random.

% robust standard errors and report these estimates in
% table~\ref{table:baltab} alongside the number of matches used to test
% for balance. Wefail to reject the null that the toss is random.

%N = 6377
%Wins = 3240 home team wins
% prob succes = 50.81
% p value = 20.14\%.

%Wins = 17189
%N = 34283
%50.14
%p value = 0.6114

%Underdog = lower ranked team in any dyad with nonmissing ranking info (so int'l matches)
%Underdog won = 2001
%N = 4054
%Win prob = 49.36\%
%p value = 0.4231

% \input{output/baltab.tex}

@@

To estimate the advantage of winning the toss, we run regressions on a team $\times$ match level dataset. Let $i$ index teams, and $j$ index matches. $Y_{ij}$ is a dummy for team $i$ winning match $j$, and takes a value of $0.5$ for draws. $Z_{ij}$ is a dummy for team $i$ winning the toss in match $j$. For the main specifications of interest, we estimate regression equations of the form
\begin{equation}\label{eqn:reduced_form}
Y_{ij} = \alpha_0 + \tau Z_{ij} + \gamma_i + \delta_t +
\psi_k + \varepsilon_{ij}
\end{equation}
which includes fixed effects for teams $\gamma_i$, time $\delta_t$, and match-type $\psi_k$, so that all variables can be interpreted as residuals from partialing out team, match-type, and time fixed effects. Since unobservable factors are correlated across opposing teams in the same match, we cluster the standard errors at the match level.

We report results in Table \ref{table:reduced_form}. We progressively introduce team, match-type, and time period fixed-effects in columns 2, 3, and 4 to make more fine-grained comparisons, and find that results are remarkably stable across specifications. The first column, for example gives
the baseline estimate of the toss effect being 1.8 percentage points in all matches. As we mentioned in the introduction, this effect represents toss winners winning 39.2\% of the time and losing 37.4\% of the time, with the remaining matches being draws. The estimated effect remains at 1.7 percentage points when we progressively add the fixed effects.

We then proceed to examine the effect of the toss withing various subgroups. We report the estimates graphically in Figures \ref{fig:rf_het_TE} - \ref{fig:omni_het_test} with point estimates and sample sizes as annotations. The results are summarized as follows.

\textbf{Format.} Toss effects are most pronounced in test matches followed by first class matches, and are statistically indistinguishable from zero in ODIs and T20s. We estimate the effect of the toss on winning a test match to be 4.7 percentage points, allowing for the fact that many tests end in draws. Given that only 65\% of test matches end with one team winning, the effect of the toss conditional on the match producing a winner is higher: 7.2 percentage points. The effect of the toss in first class matches is 2.3 percentage points. These results support the conventional wisdom among lay cricket followers that the toss grants the greatest advantage in multi-day affairs. Pitches invariably deteriorate over the course of these matches, and the fourth innings in test match is often the most challenging time to bat. The pitch deteriorates far less over the course of a day, or in the case of T20s, a few hours. Our null results for limited over matches also extend \citet{de1998winning}, who, based on analysis of data from 427 international one-day matches conclude that ``winning the toss at the outset of a match provides no competitive advantage.''

\textbf{International vs.~Domestic matches.} Toss effects are virtually the same for international and domestic matches (point estimates are 1.7 percentage points in the former, and 1.9 percentage points in the latter). This is a bit surprising given differential use of cricket formats differs in international and domestic matches (for example, there are no domestic tests), and our findings of heterogeneous effects across formats.

\textbf{Effects over time.} Toss effects overall appear to have declined over time from a high of 9.9 percentage points in the second half of the 19th century, to 1.4 percentage points in the last decade. We also look at these trends within format. T20s have not been played long enough for us to draw any conclusions about how toss effects have changed over time. But within the other formats, we find that toss effects follow a similar pattern, except that in test cricket there is a peculiar uptick in in the last decade: the toss effect jumps from 0 in the decade before to 14.5 percentage points now.  We investigate potential sources of this variation below  but there are some factors that could be important drivers for which we do not have data. For example, cricketers may have become more fit, and able to play for longer periods without tiring as much as they once did. Pitches now may be more resilient to wear than they once were. Teams may have acquired better knowledge on how to play more effectively whether they bat first or second. Etc.

\textbf{Relative strength.} Common sense suggests that the toss should be less important in lopsided matches where there is one dominant team that is expected to win, whether it bats first or second. It should be more important in competitive matches where the opposing teams are more even in cricketing strength. We define competitive matches as those in which the gap in ELO ratings of the two teams is less than 15 points. Since the median ELO rating difference is 17 points, this threshold roughly splits our sample of matches for which we have ELO ratings for both teams in two halves. Surprisingly, we find the toss effect to be nearly the same, and indistinguishable from zero within both competitive and noncompetitive matches even when we look at tests and international ODIs separately (these are the two formats for which ELO ratings apply). Part of the reason for this is that the ELO ratings are available only for recent years, and we saw above that toss effects have declined over time. We also perform an omnibus test of treatment effect heterogeneity by estimating sorted partial effects \citep{chernozhukov2018sorted,chen2019sortedeffects} with respect to ranking and result history, and find minimal evidence for effect heterogeneity.

\textbf{Continents.} We find that toss effects vary only somewhat across continents, hovering around 2.5 percentage points for matches played in Asia, Oceania, and the Americas (though the estimates for the Americas are noisier in part due to the smaller number of matches played there). In Europe (mainly England) the toss effect is only 1.1 percentage points (though we do not rule out a null effect), and in Africa it is essentially zero. Overall, despite there being much talk about the different nature of wickets across the world, with wickets in certain regions being drier, or stickier, or subject to more wear, these differences do not seem to translate into substantial toss effect differences at least when we look coarsely at averages.

\textbf{Day/Night matches.} It is often claimed that the toss is more crucial in day-and-night matches. We find some evidence for this. In the limited overs formats where day/night matches take place (ODIs and T20s), the advantage of winning the toss in a day/night match is 3.3 percentage points, whereas the
advantage of winning the toss in a limited overs match played during the day is zero.

\textbf{Seasonality.} Weather is also thought to have a large impact on the playing conditions in cricket. Cool
overcast weather, for example, is thought to aid swing bowling, especially in English pitches. More generally,
the advantage of winning the toss likely varies by the weather. Although do not have data on
the weather, we can proxy it with seasons. In particular, students of the game suspect that
the advantage of winning the toss in the early English season is especially significant.
There is some evidence of a seasonal pattern for matches played in England, with the advantage of winning the toss
growing over the summer, though the trend is weak and broken in the month of July.

\textbf{Rain delays.} Finally, we look at the differential effects of winning the toss in ODI matches that were rain-delayed, causing the number of overs bowled in the second innings to be adjusted down. We examine the subset of matches where rain-delays result in revised targets computed using the \citet{duckworth1998} method. This method has been the subject of much controversy, with many cricket fans suggesting that it is rigged against the chasing team. To our surprise, we find no evidence that toss winners are able to exploit the possibility of second inning rain delays to their advantage. If anything, we see that the toss-winner has a \emph{disadvantage} of 2.4 percentage points in matches subject to the Duckworth-Lewis adjustment, though this point estimate is statistically indistinguishable from zero \footnote{Toss-winners won 160 out of 333 ODIs with DL adjustment and 3085 out of 6187 ODIs without DL adjustment}. This, however, does not mean that the Duckworth-Lewis method is not rigged against the chasing team; that issue will be discussed further in the next section.

% \begin{table}
% \centering
% \caption{Reduced Form Effect of Toss on win probability by match type}
% \input{output/reduced_form_het.tex}
% \label{table:reduced_form_het}
% \end{table}

@@

% \newpage

@@

\begin{figure}
    \centering
    \includegraphics[width=0.5\textwidth]{output/het_rank_te_odi.pdf}
    \caption{Treatment effect heterogeneity by ranking, toss and win history, and team dummies. Minimal evidence of effect heterogeneity}
    \label{fig:omni_het_odi}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[width=0.5\textwidth]{output/het_rank_te_test.pdf}
    \caption{Treatment effect heterogeneity for test matches by ranking, toss and win history, and team dummies. Minimal evidence of effect heterogeneity}
    \label{fig:omni_het_test}
\end{figure}

>>>>

\title{Asymmetries in Cricket: The Effects of Winning the Toss on Winning the Match\thanks{This paper incorporates some of the text and results from Sood \& Willis's (2019) working paper: ``Fairly Random: The Impact of Winning the Toss on the Probability of Winning.'' The authors thank Matt Blackwell, Andrew Gelman, Donald Green, Jens Hainmueller, and Daniel Stone for their comments.}}

@@

\section{Introduction}

In competitive sports, every little thing matters. Nevertheless, many sports leave some key levers out of teams' control and in the hands of fate. In cricket, one such lever---the toss---has been the subject of much attention. In many matches, commentators and cricket enthusiasts claim that there is a definite advantage to either bowling or batting first.\footnote{See, for example, the \href{https://www.quora.com/How-much-does-a-toss-matter-to-a-game-of-cricket}{Quora page on ``How much does a toss matter to a game of cricket?''}} The advantage is also often pointed to by the captains in the pre-toss interview and by the captain of the losing team in the post-match interview. To wrest the advantage, the captain merely needs to pick the side of the coin that will be left facing the sky after the toss and choose whichever one---batting first or batting second---gives the greater advantage.

In this paper, we measure how well teams have been able to capitalize on any advantage conferred by winning the toss. To answer the question, we collected data on over 35,000 cricket matches across four formats (First Class, ODIs, T20s, and Tests) played over the course of more than one hundred fifty years.

We find that winning the toss increases the chances of winning the match by 1.8 percentage points: toss winners prevail 39.2\% of the time and lose 37.4\% of the time; the rest of the matches are drawn or tied. The effect of winning the toss is the largest in Tests, where it is 4.9 percentage points and the smallest in ODIs and T20s, where it is essentially zero on average. That said, in day/night ODIs and T20s, the effect of winning the toss is 3.3 percentage points; in day-time matches, the effect is a precise zero.

The effect of the toss in ODIs has remained relatively stable since the 1980's when the one-day format first became popular. In Tests, however, the importance of the toss has varied considerably. The effect of winning the toss in Test cricket was a little over five percentage points before 1990. It then dropped to almost 0 from 1990 to 2010. In the last decade, however, it has exploded, averaging 14.5 percentage points.

Contrary to conventional wisdom, the effect of winning the toss does not vary dramatically across continents, though the toss appears less critical on European and African wickets than elsewhere. The effect on Aussie and Kiwi wickets (3 percentage points) is not substantially different from its effect on South Asian wickets (2.8 percentage points) but is higher than on European and African wickets (1.1 and 0 percentage points respectively).

@@

However, we find that of the toss winners that elected to bat, only 38.1\% won the match compared to 37.9\% that lost--- a small difference of 0.2 percentage points. On the other hand, of the toss winners that elected to field, 40.8\% won the match compared to 36.9\% that lost--- a larger difference of 3.9 percentage points. This means that 94\% of the 1.8 percentage point effect of winning the toss can be attributed to instances in which the toss winning team elected to field, and only 6\% of the effect can be attributed to the toss winning team electing to bat. This suggests that teams are better able to capitalize on the toss by electing to field when it is advantageous to do so.

We also look at how the toss advantage breaks down between teams that elected to bat versus those that elected to field within the various subgroups mentioned above and others. Perhaps the most striking result from this analysis for many cricket followers is that there does not appear to be the case that toss winners that elect to bat first do better than those that elect to field first in rain delayed matches where the chasing total is adjusted according to the \citet{duckworth1998} formula.

Lastly, a caveat about our findings. Our empirical approach estimates the extent to which teams have been able to capitalize on the advantage granted by winning the toss. The toss may matter a lot in principle, but teams may decide poorly. So, when we refer to ``toss effects'' in this paper, we do not refer to whether there is an advantage to winning the toss in principle, whether any such advantage is exploited in practice.

Our results are pertinent to debates about how cricket should be governed. One argument is that since the toss is random, no team is advantaged by the format prior to the toss; and in fact the luck of the toss along with the strategic choice to bat first or second are integral parts of the game--- not very different from the luck of weather affecting the teams differently.

An alternative argument, however, is that the toss is problematic because it advantages one team over the other even before the first ball is bowled; and care should be taken to mitigate this asymmetry--- for example, by making cricket pitches more durable so that they don't deteriorate so much so that batting on the final day of a test match is miserable. Otherwise, the toss-winning team could exploit this advantage by batting first, reducing the chance that it has to endure this misery. Moreover, the fact that no side is advantaged \emph{ex-ante} if the toss is random is not enough to justify not correcting any asymmetry created by the toss. In fact, the rules of cricket already try to correct the asymmetry between teams that bat first and those that bat second in one particular instance: In rain-delayed one-day matches, the Duckworth-Lewis formula is used to set a new target for the chasing team.\footnote{The argument that no side was advantaged \emph{ex-ante} could have been used to justify not doing any correction in such rain-delayed matches, and simply declaring that the team that batted first won the match because the chasing team was unable to reach its target. But the designers of one-day cricket were clearly unwilling to allow the bad luck of weather to have this much influence on the game. If addressing the misfortunes of weather are important, it may also be important to address the misfortune of losing the toss.}

@@

We do not address these optimal design questions in this paper, but our results are relevant to these issues depending on what the designers of cricket seek to optimize. However, at the end of the paper, we propose (and discuss the possible consequences of) a subtle change to the rules governing the toss: Rather than tossing a coin to confer the choice to bat first or second upon the captain of the toss-winning team, umpires may consider simply tossing a coin to determine which team must bat first.

@@

We obtained data on cricket matches from the ESPNCricinfo website, which covers international and domestic matches, and maintains statistics on a wide variety of matches across formats.\footnote{Information on the matches was downloaded in June 2020 using a purpose-built library available at \href{https://github.com/dwillis/python-espncricinfo}{https://github.com/dwillis/python-espncricinfo}. The data is available at \href{https://github.com/dwillis/cricket-stats}{https://github.com/dwillis/cricket-stats}.}

We focus on four formats: T20, ODI, Test, and First Class for both men's and women's teams.\footnote{We exclude List A matches that are not ODIs or T20s. These matches are considered a less professional category (with worse data recording) than the formats that we include.} The data are a near census of the relevant population, only excluding scheduled matches that were abandoned without the toss being conducted, and a small number of limited-overs matches in which the minimum number of overs that must be bowled to establish a result were not met.\footnote{In ODIs, for example, each side must bat at least 20 overs for a result to be declared. We have no reason to suspect that the small matches that were excluded because they did not meet this threshold could be biased enough to affect our conclusions. We drop 426 matches recorded as ``no result,'' 3 matches that are recorded as ``abandoned'' or ``canceled,'' and another three that don't record who batted first. The "no result" matches constitute the largest exclusion we make. They appear to be comprised of matches canceled for idiosyncratic reasons, ranging from the \href{https://www.espncricinfo.com/series/shell-tri-series-1991-92-61203/australia-women-vs-england-women-final-66975/full-scorecard}{a team being unable to bat because of the weather} to cancellations because of \href{https://www.espncricinfo.com/series/australia-tour-of-pakistan-1982-83-61392/pakistan-vs-australia-3rd-odi-64196/full-scorecard}{unwanted audience participation}. In all, we drop approximately 0.07\% of matches.}

Our final sample comprises 35,357 cricket matches across four formats played over the last 150 years. We report the distribution of matches over time and by type in Figure \ref{fig:timeseries_decomp}. As the figure shows, the number of matches has exploded over the last two decades, largely due to the introduction of the enormously popular Twenty-20 (T20) format of the game both in international and domestic tournaments. T20 matches constitute over a quarter of all matches played in the last decade.

@@

Since our analysis rests on the assumption that the coin toss is random, we start by examining this assumption.\footnote{Prior work by \citet{shafqat2015analysis} and \citet{dawson2009bat} conditions on post-treatment variables, whereas we use only the randomization of the toss to identify toss effects.} Is there any evidence that the coin toss is rigged---for example, does the home side enjoy the rub of the green more often? In order to study this, we compare the rates at which home teams, teams that won the toss in their last match, and underdogs (defined as the lower-ranked team whenever rankings are available for both teams) win the toss relative to their opponents.

To examine whether the toss is rigged, we compute the odds of seeing as many successes in the same number of draws of a sequence of independent Bernoulli random variables with parameter 0.5. For example, home teams have won 50.81\% of tosses in the 6,377 matches for which we have data. The likelihood of tossing a fair coin 6,377 times and getting at least this percentage of heads is 20.1\%. This chance appears low, but not eyebrow-raisingly so. We find that toss wins are even more balanced for teams that won and lost the last toss, and for teams that are underdogs and favorites. Table \ref{table: balance} reports these balance results.

That said, we do see one potential source of concern. We find that teams that share their nationality with at least two of the three umpires of the match win 51.3\% tosses in the 4,044 international matches for which we have this data. The likelihood of tossing a fair coin 4,044 times and getting at least this share of heads is only 10\%. This number is certainly low enough to raise some eyebrows. That said, the practice of allowing home-team umpires in international matches ended in 1992, and so these issues are less relevant in recent decades which comprise the bulk of our sample. In addition, a rigged toss in favor of the home team would bias estimates of the advantage of winning the coin toss to the extent that it is correlated with ability. If stronger teams win more tosses, estimates of the advantage of the coin toss would be inflated upwards. And vice versa, if otherwise. Given home teams win more matches and are seen as 'stronger' at home, it is possible that the results are slightly biased upwards. We investigate this point later in the paper, looking at effects post 1992. For now, we proceed as if the toss was random.

@@

To estimate the advantage of winning the toss, we run regressions on a team $\times$ match level dataset. Let $i$ index teams, and $j$ index matches. $Y_{ij}$ is a dummy for team $i$ winning match $j$, and takes a value of $0.5$ for draws. $Z_{ij}$ is a dummy for team $i$ winning the toss in match $j$. For the main specifications of interest, we estimate regression equations of the form

\begin{equation}\label{eqn:reduced_form}
Y_{ij} = \alpha_0 + \tau Z_{ij} + \gamma_i + \delta_t + \psi_k + \varepsilon_{ij}
\end{equation}

which include fixed effects for teams $\gamma_i$, time $\delta_t$, and match-type $\psi_k$. We cluster the standard errors at the match level.

We report results in Table \ref{table:reduced_form}. We progressively introduce team, match-type, and time period fixed effects in columns 2, 3, and 4, and find that results are remarkably stable. The first column gives the baseline estimate of the toss effect. The 1.8 percentage points difference reflects the fact that toss winners win 39.2\% of the time and lose 37.4\% of the time; the remaining matches are drawn. The estimated effect remains at 1.7 percentage points when we progressively add various fixed effects.

We then proceed to examine the effect of the toss within various subgroups. We report the estimates graphically in Figures \ref{fig:rf_het_TE} - 4 with point estimates and sample sizes as annotations. We summarize the results below:

\textbf{Format.} Toss effects are most pronounced in Test matches followed by first class matches, and are statistically indistinguishable from zero in ODIs and T20s. We estimate the effect of the toss on winning a Test match to be 4.7 percentage points. The effect of the toss in first class matches is 2.3 percentage points. These results support the conventional wisdom among lay cricket followers that the toss grants the greatest advantage in multi-day affairs. Pitches invariably deteriorate over the course of these matches, and the fourth innings in a Test match is often the most challenging time to bat. The pitch deteriorates far less over the course of a day, or in the case of T20s, a few hours. Our null results for limited over matches also extend \citet{de1998winning}, who, based on analysis of data from 427 international one-day matches conclude that ``winning the toss at the outset of a match provides no competitive advantage.''

\textbf{International vs.~Domestic matches.} Toss effects are virtually the same for international and domestic matches (point estimates are 1.7 percentage points in the former, and 1.9 percentage points in the latter).

\textbf{Effects over time.} Toss effects appear to have declined over time from a high of 9.9 percentage points in the second half of the 19th century to 1.4 percentage points in the last decade. This trend may be a result of a changing composition of matches so we also looked at these trends within formats. T20s have not been played long enough for us to draw any conclusions about how toss effects have changed over time. But within the other formats, we find that toss effects follow a similar pattern, except that in Test cricket there is a peculiar uptick in the last decade: the toss effect jumps from 0 in the decade before to 14.5 percentage points now. We investigate potential sources of the ecological decline below but there are some factors that could be important drivers but for which we do not have the data. For example, cricketers may have become fitter, and able to play for longer periods without tiring as much as they once did, reducing the impact of the order in which they ball or bat. Pitches may have become more resilient to wear than they once were. Teams may have become better informed about how to play more effectively whether they bat first or second, etc.

\textbf{Relative strength.} We expect the toss to be less important in matches where one team is clearly stronger than the other. We expect the stronger team to win, whether it bats first or second. To test this hypothesis, we use team ranking data published by cricket's official governing body, the International Cricket Council (ICC). The median match is played between teams where the ICC rating difference is 17 points. We use that as the approximate splitting point, splitting the matches into teams where the gap in the ICC ratings of the two teams is less than 15 points and where it is more. In this case, we believe the standard errors are too large for us to arrive at any substantively meaningful conclusions. Statistically, we believe effects larger than 8\% are unlikely.

\textbf{Continents.} Toss effects vary only somewhat across continents, hovering around 2.5 percentage points for matches played in Asia, Oceania, and the Americas (though the estimates for the Americas are noisier in part due to the smaller number of matches played there). In Europe (effectively England) the toss effect is only 1.1 percentage points (though we cannot rule out a null effect), and in Africa it is essentially zero. Overall, despite there being much talk about the different kinds of wickets across the world, with wickets in certain regions being drier, or stickier, or subject to more wear, these differences do not seem to translate into substantial toss effect differences at least when we look at coarse geographic averages.

\textbf{Day/Night matches.} It is often claimed that the toss is more crucial in day-and-night matches. We find some evidence for this. In the limited overs formats where day/night matches take place (ODIs and T20s), the advantage of winning the toss in a day/night match is 3.3 percentage points, whereas the

advantage of winning the toss in a limited overs match played during the day is zero.

\textbf{Seasonality.} Weather is also thought to have a large impact on the playing conditions in cricket. Cool overcast weather, for example, is thought to aid swing bowling, especially on English pitches. More generally, the advantage of winning the toss likely varies by weather. Although we do not have data on the weather, we can proxy it with seasons. In particular, students of the game suspect that the advantage of winning the toss in the early English season is especially significant. There is some evidence of a seasonal pattern for matches played in England, with the advantage of winning the toss growing over the summer, though the trend is weak and broken in the month of July.

\textbf{Rain delays.} Finally, we look at the differential effects of winning the toss in ODI matches that were rain-delayed, causing the number of overs bowled in the second innings to be adjusted down. We examine the subset of matches where rain-delays result in revised targets computed using the \citet{duckworth1998} method. This method has been the subject of much controversy, with many cricket fans suggesting that it is rigged against the chasing team. To our surprise, we find no evidence that toss winners are able to exploit the possibility of second-inning rain delays to their advantage. If anything, we see that the toss-winner has a \emph{disadvantage} of 2.4 percentage points in matches subject to the Duckworth-Lewis adjustment, though this point estimate is statistically indistinguishable from zero.\footnote{Toss-winners won 160 out of 333 ODIs with the Duckworth-Lewis adjustment (i.e.~48\%) and 3085 out of 6187 ODIs without the Duckworth-Lewis adjustment (49.9\%).} This, however, does not mean that the Duckworth-Lewis method is not rigged against the chasing team; that issue will be discussed further in the next section.

@@

@@
